---
title: "Project443"
author: "Shahrukh Islam Prithibi (47610068)"
date: "2026-04-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we will import libraries and set seeds.

```{r import_data, message=FALSE, warning=FALSE}
library(tidyverse)
library(forecast)
set.seed(443)
```

Next, appropriately rename labels and set date column correctly.

```{r rename_labels, warning=FALSE}
dataset <- read.csv(file = "https://raw.githubusercontent.com/ShahrukhP15/Time-Series-Forecasting-of-U.S.-Fuel-Price-Spreads/refs/heads/main/PET_PRI_GND_DCUS_NUS_W.csv")
new_col_names <- c( 
  all_grades_all_forms = "A1", all_grades_conv = "A2", all_grades_reform = "A3",
  regular_all_forms = "R1", regular_conv = "R2", regular_reform = "R3", 
  midgrade_all_forms = "M1", midgrade_conv = "M2", midgrade_reform = "M3",
  premium_all_forms = "P1", premium_conv = "P2", premium_reform = "P3",
  diesel_no2 = "D1")

dataset <- dataset |>
  rename(new_col_names) |>
  mutate(Date = mdy(Date)) |> 
  arrange(Date) |> 
  drop_na()

head(dataset)
```

Check if there are any gaps in the weekly data.

``` {r data_checks}
date_diffs <- dataset |>
  mutate(diff = as.numeric(Date - lag(Date))) |>
  filter(!is.na(diff))

print(min(date_diffs$diff) - max(date_diffs$diff)) # If 0, then all values are equal, as needed.
```

Plot the data.

```{r preliminary_plots}
dataset |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "price") |>
  ggplot(aes(x = Date, y = price, color = type)) +
  geom_line(alpha = 0.7) +
  labs(title = "US Gasoline Prices Over Time",
       x = "Date", y = "Price ($/gallon)", color = "Fuel Type") +
  theme_minimal()
```
A lot of the lines seem very close. It may be worth comparing the differences of them
to the "average", which we will use `all_grades_all_forms` for.

```{r vs_average}
dataset_centered <- dataset |>
  mutate(across(-Date, ~ . - all_grades_all_forms))

dataset_centered |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "price") |>
  ggplot(aes(x = Date, y = price, color = type)) +
  geom_line(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Gas Prices Relative to All Grades All Forms",
       x = "Date", y = "Price Difference ($/gallon)", color = "Fuel Type") +
  theme_minimal()
```

Here, we can see that diesel stands out a lot compared to the rest. There seems
to be exponential growth between the other values. Let's do a log ratio plot.

```{r log_ratio_plot}
dataset_log <- dataset |>
  mutate(across(-Date, ~ log(. / all_grades_all_forms)))

dataset_log |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "log_ratio") |>
  ggplot(aes(x = Date, y = log_ratio, color = type)) +
  geom_line(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Gas Prices Log Ratio Relative to All Grades All Forms",
       x = "Date", y = "Log Price Ratio", color = "Fuel Type") +
  theme_minimal()
```

Next, we will see how each one varies by gas grade.

```{r gas_by_grade}
dataset_log |>
  select(-all_grades_all_forms) |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "log_ratio") |>
  mutate(grade = case_when(
    str_detect(type, "regular")    ~ "Regular",
    str_detect(type, "midgrade")   ~ "Midgrade",
    str_detect(type, "premium")    ~ "Premium",
    str_detect(type, "all_grades") ~ "All Grades"
  )) |>
  filter(grade != "All Grades") |>
  ggplot(aes(x = Date, y = log_ratio, color = grade, group = type)) +
  geom_line(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "By Grade", x = "Date", y = "Log Price Ratio", color = "Grade") +
  theme_minimal()
```

Similarly, we will see how each plot varies by formulation.

```{r gas_by_formulation}
dataset_log |>
  select(-all_grades_all_forms) |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "log_ratio") |>
  mutate(formulation = case_when(
    str_detect(type, "conv")   ~ "Conventional",
    str_detect(type, "reform") ~ "Reformulated",
    TRUE                       ~ "All Forms"
  )) |>
  filter(formulation != "All Forms") |>
  ggplot(aes(x = Date, y = log_ratio, color = formulation, group = type)) +
  geom_line(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "By Formulation", x = "Date", y = "Log Price Ratio", color = "Formulation") +
  theme_minimal()
```

Lowkey, a lot of the graphs above can go into the appendix instead.

In this paper, we want to find differences between diesel and gasoline. Let's plot some of these differences. First, modify the dataset to show the difference between diesel and all the other gasoline.

```{r diesel_gas_diff}
dataset_diff_diesel <- dataset |>
  mutate(across(-c(Date), ~ . - diesel_no2))

```


Next, we will plot the differences.

```{r diesel_gas_plot}
dataset_diff_diesel |>
  select(-diesel_no2) |>
  pivot_longer(cols = -Date, names_to = "type", values_to = "diff") |>
  ggplot(aes(x = Date, y = diff, color = type)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Gasoline Prices Relative to Diesel Over Time",
       x = "Date", y = "Price − Diesel ($/gallon)", color = "Fuel Type") +
  theme_minimal()
```

It looks like the different types of gas follow the same distribution, with some slight variance within each other. Either way, they are highly correlated. It will be good to focus on one type of gas instead of all of them, because they all show similar patterns.

We will focus on All Grades All Forms, as it is a good baseline of all the different types of gas. We can plot this more clearly.

```{r focused}
dataset_focused <- dataset |>
  select(Date, diesel_no2, all_grades_all_forms) |>
  mutate(diff = all_grades_all_forms - diesel_no2)


dataset_focused |>
  ggplot(aes(x = Date, y = diff)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Gasoline vs. Diesel Price Difference Over Time",
       x = "Date", y = "All Grades All Forms − Diesel ($/gallon)") +
  theme_minimal()
```

Let's see what the graphs look like when logging it.

```{r log_diffs}
dataset_focused <- dataset_focused |>
  mutate(
    log_gas = log(all_grades_all_forms),
    log_diesel = log(diesel_no2),
    log_ratio = log_gas - log_diesel
  )


dataset_focused |>
  ggplot(aes(x = Date, y = log_ratio)) +
  geom_line(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Log Ratio of Gasoline to Diesel Over Time",
       x = "Date", y = "Log(Gas / Diesel)") +
  theme_minimal()
```


Based on the graph above, it seems that the price difference between diesel and gas were somewhat stable before around 2008. However, since then, the price of gas has been consistently lower than diesel. It may be worth splitting our dataset up into before and after 2008 to see if the time series analysis that we do on them differs between these frames. There also seems to be somewhat regular shifts between spikes. This may indicate a level of seasonality. We will explore this more later. 


Below we have a summary statistics table 

```{r summary_statistics}
summary_statistics <- dataset_focused |>
  select(all_grades_all_forms, diesel_no2, diff, log_ratio) |>
  pivot_longer(everything(), names_to = "type", values_to = "value") |>
  group_by(type) |>
  summarise(
    mean      = mean(value),
    abs_mean  = mean(abs(value)),
    sd        = sd(value),
    variance  = var(value),
    min       = min(value),
    max       = max(value)
  )

summary_statistics
```

Next, we have ACF plots with the difference and log ratio

```{r acf, warning=FALSE}
gridExtra::grid.arrange(
  ggAcf(dataset_focused$diff,      main = "ACF of Price Difference"),
  ggAcf(dataset_focused$log_ratio, main = "ACF of Log Ratio"),
  ncol = 2
)
```

From this graph we can interpret a few things. Firstly, it seems that taking the log ratio
does not seem to make much of a difference, as the ACF plot looks nearly the same. Second, because all the ACF values are large throughout, it indicates that the data may not be stationary. We will likely need to do some differencing to achieve stationarity. 

```{r diffdiff, warning=FALSE}
dataset_focused <- dataset_focused |>
  mutate(diff_log_ratio = c(NA, diff(log_ratio)))

ggAcf(na.omit(dataset_focused$diff_log_ratio), main = "ACF of Differenced Log Ratio")

```

This looks much better. We see that the ACF drops to 0 quite quickly now. Also, we see that there is a negative spike at round 26 weeks (or half a year), which may indicate some seasonality in the data. 